# Opportunistic data on 🐲-flies

### Name searching

In this example we are interested in exploring opportunistically collected data from the Swedish citizen science species observation portal - Artportalen.

To begin with, we want be sure there is an unequivocal way to find the species within the order Odonata (dragonflies) and nothing else, so let's search for "odonata":

```{r error=FALSE, message=FALSE, warning=FALSE, include=FALSE}
library(SBDI4R)
library(sf)
library(BIRDS)
```

```{r search1, echo=TRUE}
sx <- search_fulltext("odonata")
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```

### 

::: {.notes}
We quickly see there that other taxonomic levels appear too, and also species that look suspiciously as not belonging to dragonflies. But there is only one order.
:::

Let's refine the search. To know which search fields we can use to filter the search we use the function `sbdi_fields(fields_type = "general")`. The search field we are looking for is **"order_s"**.

```{r search2, echo=TRUE}
sx <- search_fulltext(fq="order_s:Odonata", page_size = 10)
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```

### 

Now we can download the taxonomic data (note that the search is case-sensitive):

```{r taxinfo, message=FALSE, results=FALSE}
tx <- taxinfo_download("order_s:Odonata", 
                       fields = c("guid", "order_s","genus_s", "specificEpithet_s", 
                                  "scientificName",  "canonicalName_s", "rank"), 
                       verbose = FALSE)
## restrict to species and not hybrids
tx <- tx[tx$rank == "species" & tx$genusS != "",] 
```

You can save the `tx` object as the complete species list for later use.

### Filters

We start by searching for the data resource we are interested in using the function `pick_filter()`. This is an interactive query guiding you through the many resources available to filtering your query (data resources, spatial layers, and curated species lists).

```{r filter2.1, message=FALSE, eval=FALSE}
# follow the instructions 
fq_str <- pick_filter("resource") 
```

Follow the instructions. Your choices here would have been "in3" ▶️ "dr5". Your variable `fq_str` will now contain a string "`data_resource_uid:dr5`".

```{r filter2.2, echo=FALSE, eval=TRUE}
fq_str <- "data_resource_uid:dr5"
```

We only want to look at data from year 2000 to 2010:

```{r filter2.3}
y1 <- 2000
y2 <- 2010
fq_str <- c(fq_str, paste0("year:[", y1, " TO ", y2,"]"))
# Note the square brackets are hard limits
```

### Spatial filter

We also want to filter spatially for Southern Sweden ([Götaland](https://en.wikipedia.org/wiki/G%C3%B6taland)).

SBDI APIs take as search input polygons in the so-called WKT [Well Known Text](https://www.geoapi.org/3.0/javadoc/org/opengis/referencing/doc-files/WKT.html) format.

```{r sweshape}
data("swe", package = "SBDI4R")
wGotaland <- swe$Counties$LnNamn %in% c("Blekinge", "Gotlands", "Hallands", "Jönköpings", "Kalmar",
                                        "Kronobergs", "Östergötlands", "Skåne", "Västra Götalands")
gotaland_c <- swe$Counties[wGotaland,]
```

::: {.notes}
Vector spatial layers (e.g.. polygons) can be imported in a number of different ways. SBDI APIs take as search input polygons in the so-called WKT [Well Known Text](https://www.geoapi.org/3.0/javadoc/org/opengis/referencing/doc-files/WKT.html) format. So the first step is to load a vector layer and transform it into a WKT string. You could instead use the data we provide in the SBDI4R package `data("swe")`.

There are details about this polygon that we need to take care before. The WKT string should not be too long to be accepted by the API service. Also, the polygon we just got is projected in the coordinate system SWEREF99 TM, and the API service only accepts coordinates in a geodesic coordinate system WGS84.

NOTE: as of today, the SBDI APIs will only work properly if the polygon is submitted as a MULTIPOLYGON
:::

Let's construct the WKT string:

```{r searchpoly2WKT, warning=FALSE}
# transform the CRS
gotaland_c <- st_transform(gotaland_c, crs = st_crs(4326))

# disolve the counties into one polygon
gotaland <- st_union(gotaland_c)

# create a convex hull of the polygon
gotaland_ch <- st_convex_hull(gotaland)

# cast it as MULTIPOLYGON as this is what SBDIs API need
gotaland_ch <- st_cast(gotaland_ch, to = "MULTIPOLYGON")

# create WKT string
wkt <- st_as_text(gotaland_ch)
```

### 

The WKT string then looks like this:

```{r, echo=FALSE}
wkt
```

```{r searchpoly, fig.width=7, echo=FALSE, warning=FALSE, fig.align='center'}
#library(sf)
oldpar <- par()
par(mar = c(0,0,0,0))
plot(gotaland_c$geometry, border="blue")
plot(gotaland, lwd=2, add=TRUE)
plot(gotaland_ch, lwd=2, border="red", add=TRUE)
legend("bottomleft", inset = c(0,0.05),
       legend = c("counties", "Götaland", "search polygon"),
       col = c("blue","black", "red"),
       lwd = c(1,2,1),
       title = NULL,bty="n")
suppressWarnings(par(oldpar))
```

### Occurrences

Next, we download the observations using the command `occurrences()`.

::: {.notes}
Next, we download the observations using the command `occurrences()`, but be aware that the search fields may not be the same as those used to search for taxa. We therefore recommend using the function `sbdi_fields("occurrence")` to find out which search fields we can use to filter for occurrences. Here we see that the field we need this time is "order".
:::

```{r getData, cache=TRUE}
xf <- SBDI4R::occurrences(taxon = "order:Odonata", 
                          fq = fq_str, 
                          ### <b>
                          wkt = wkt, 
                          ### </b>
                          extra = "collector")
```

```{r localtmp, echo=FALSE, eval=FALSE}
save(xf, file = "data/tmp_occdata2_for compile.rdata")
load(file = "data/tmp_occdata2_for compile.rdata")
```

```{r occdata, include=FALSE}
head(xf$data)
```

### Quality and fit-for-use check

::: {.notes}
Before we can use the observation records we need to know if the observation effort (sampling effort) has varied over time and in space. We can approximate observation effort from the data by defining field visits i.e. occasions at which an observer has sampled observations. We reconstruct field visits (that is, assign each observation a visitUID) using the package [BIRDS](https://greensway.github.io/BIRDS/). Additionally we want the data to be summarized over a grid of 25 km (provided through the SBDI4R package). The following functions will perform many different summaries at the same time. Please refer to the BIRDS package documentation for more detail.
:::

Before we can use the observation records we need to know if the observation effort (sampling effort) has varied over time and in space. we can

We can approximate observation effort from the data by defining field visits i.e. occasions at which an observer has sampled observations using the package [BIRDS](https://greensway.github.io/BIRDS/). Additionally we want the data to be summarized over a grid of 25 km. Please refer to the BIRDS package documentation for more detail.

```{r birds, message=FALSE, warning=FALSE, eval=FALSE}
remotes::install_github("Greensway/BIRDS")
library(BIRDS)
```

### Organise and Summarise

```{r organisebirds, message=FALSE, eval=FALSE}
OB <- organiseBirds(xf$data, sppCol = "species" , 
                    # We only want observations identified at the species level
                    taxonRankCol = "rank", taxonRank = "species", 
                    # the visits are defined by collector and named locality
                    idCols = c("locality", "collector"), 
                    timeCols = c("year", "month", "day"), 
                    xyCols =c("longitude","latitude") )
```

We don't need the whole grid, just the piece that overlaps our searching polygon

```{r thegrid, message=FALSE, eval=FALSE}
wInt <- unlist(st_intersects(gotaland, Sweden_Grid_25km_Wgs84))
gotaland_grid25 <- Sweden_Grid_25km_Wgs84[wInt,]
```

```{r summariseBirds, message=FALSE, eval=FALSE}
SB <- summariseBirds(OB, 
                     grid = gotaland_grid25, 
                     spillOver = "unique")
```

```{r save_birds, eval=FALSE, include=FALSE}
save(list=c("OB", "SB"), file = "output/BIRDS_summary.rdata")
```

```{r load_birds, include=FALSE}
load(file = "output/BIRDS_summary.rdata")
```

### 

<div>

Once summarised, we can see over space and for a few selected years how the number of observations is distributed

</div>

```{r plotBIRDSspatial, echo=FALSE, fig.width=8, fig.align='center'}
maxC <- max(SB$spatial$nObs, na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")
oldpar <- par()
par(mar = c(1,1,1,1), mfrow=c(1,3))
plot(SB$spatial$geometry, col=palBW(SB$spatial$nObs),
     border = "grey", main="All years")
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       title = "Number of \nobservations", pch = 15, bty="n")

## or export other combinations, e.g. one map per observed year
yearlySp <- exportBirds(SB, 
                        dimension = "spatial", 
                        timeRes = "yearly", 
                        variable = "nObs", 
                        method = "sum")

maxC <- max(yearlySp$'2005', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp$geometry, col=palBW(yearlySp$'2005'), 
     border = "grey",main="2005")
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")

maxC <- max(yearlySp$'2010', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp$geometry, col=palBW(yearlySp$'2010'), 
     border = "grey",main="2010")
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")
suppressWarnings(par(oldpar))
```

### 

We now want to use the number of field visits as the measure for sampling effort:

```{r ggplot1code, eval=FALSE}
vis <- ggplot(data = SB$spatial, aes( fill = nVis)) +
  geom_sf() +
  ggtitle("Number of visits")

spp <- ggplot(data = SB$spatial ,aes( fill = nSpp)) +
  geom_sf() +
  ggtitle("Number of species")
```

```{r ggplot1, echo=FALSE, fig.align='center', fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
library(cowplot)
library(ggplot2)
library(colorRamps)
library(gridExtra)

vis <- ggplot(data = SB$spatial, aes( fill = nVis)) +
  geom_sf() +
  ggtitle("Number of visits") +
  scale_fill_gradient(low = "#56B1F7",
                      high = "#132B43",
                      na.value = NA) +
  theme(plot.margin = margin(1, 1, 1, 1, "pt")) +
  theme_cowplot()

spp <- ggplot(data = SB$spatial ,aes( fill = nSpp)) +
  geom_sf()+
  ggtitle("Number of species")+
  scale_fill_gradient(low = "#56B1F7",
                      high = "#132B43",
                      na.value = NA) +
  theme(plot.margin = margin(1, 1, 1, 1, "pt")) +
  theme_cowplot()

grid.arrange(vis, spp, ncol=2)
```

### Temporal check

We see that `SB` contains an element called `SB$temporal` that contains a daily time series with time-specific rows when there is information. `xts` also supports day time, but dating below day resolution is not yet implemented in the `BIRDS` package.

```{r temporal}
sb.xts <- SB$temporal
head(sb.xts, 5)
```

Sub-setting is convenient in `xts` as you can do it with its dates and with a `/` for a range of dates.

```{r subsetting, eval=FALSE}
sb.xts["2010-09-07"] #a specific day
sb.xts["2010-09-01/2010-09-15"] #for a period
sb.xts["2010-09"] #a specific month
```

### 

The package `xts` has several tools for converting to different time periods. Here we use `apply.monthly` to obtain the total number of observations and visits per month. Read more in `?plot.xts`.

```{r monthlyPlotcode, eval=FALSE}
library(xts)
obs.m <- apply.monthly(sb.xts$nObs, "sum", na.rm = TRUE)
vis.m <- apply.monthly(sb.xts$nVis, "sum", na.rm = TRUE)

plot(obs.m, col = "darkblue", grid.ticks.on = "month", 
     major.ticks = "year", grid.col = "lightgrey",  
     main = "Total number of daily observations and visits per month")
lines(vis.m, col = "orange", lwd=2, on=1)
```

```{r monthlyPlot, fig.align='center', fig.height=3.5, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(xts)
obs.m <- apply.monthly(sb.xts$nObs, "sum", na.rm = TRUE)
vis.m <- apply.monthly(sb.xts$nVis, "sum", na.rm = TRUE)

plot(obs.m, 
     col = "darkblue", grid.ticks.on = "month", 
     major.ticks = "year", grid.col = "lightgrey",  
     main = "Total number of daily observations and visits per month")

lines(vis.m, col = "orange", lwd=2, on=1)

addLegend("topleft", c("Observations", "Visits"), 
          col=c("darkblue","orange"), 
          lwd=2)
```

```{r monthlyPlotDo, echo=FALSE, fig.align='center', fig.height=3.5, message=FALSE, warning=FALSE, paged.print=FALSE}
addLegend("topleft", c("Observations", "Visits"), 
          col=c("darkblue","orange"), 
          lwd=2)

```

### Species trends

We can now look at some particular species and ask whether those have changed in occurrence over time:

```{r sppTrendsCode, eval=FALSE}
speciesSummary(SB)
```

```{r sppTrends, echo=FALSE, message=FALSE}
speciesSummary(SB)[1:15,1:4]
cat("...")
```

### 

We pick two species and compare their trends in number of visits where the species where reported, relative to the total number of visits.

```{r trends, eval=FALSE}
library(dplyr)
sppCount <- obsData(OB) |> 
    group_by(year, visitUID) |> 
    summarise("focalCountLq" = sum(scientificName == "Libellula quadrimaculata"),
              "focalCountSd" = sum(scientificName == "Sympetrum sanguineum"),
              "sppLength" = length(unique(scientificName)), 
              .groups = "drop") |> 
    ungroup() |> 
    group_by(year) |> 
    summarise("focalCountLq" = sum(focalCountLq),
              "focalCountSd" = sum(focalCountSd),
              "nVis" = length(unique(visitUID)),
              "relCountLq" = focalCountLq / nVis,
              "relCountSd" = focalCountSd / nVis,
              .groups = NULL)

plot(sppCount$year, sppCount$relCountLq, 
     type = "l", lwd = 3, xlab = "Year", 
     ylab = "Relative number of visits with observations", 
     ylim=c(0, max(sppCount$relCountLq)), 
     xaxp=c(2000, 2010, 10))
lines(sppCount$year, sppCount$relCountSd, lwd=3, col="#78D2EB")
legend("bottomright", 
       legend=c("Libellula quadrimaculata","Sympetrum sanguineum"), 
       text.font = 3, col = c("black", "#78D2EB"), lwd = 3, bty = "n")
```

### 

```{r trendsplot, echo=FALSE, fig.width=9, fig.height=6, fig.align='center'}
library(dplyr)
sppCount <- obsData(OB) |> 
    group_by(year, visitUID) |> 
    summarise("focalCountLq" = sum(scientificName == "Libellula quadrimaculata"),
              "focalCountSd" = sum(scientificName == "Sympetrum sanguineum"),
              "sppLength" = length(unique(scientificName)), 
              .groups = "drop") |> 
    ungroup() |> 
    group_by(year) |> 
    summarise("focalCountLq" = sum(focalCountLq),
              "focalCountSd" = sum(focalCountSd),
              "nVis" = length(unique(visitUID)),
              "relCountLq" = focalCountLq / nVis,
              "relCountSd" = focalCountSd / nVis,
              .groups = NULL)


oldpar <- par(no.readonly = TRUE)
par(mar=c(4,4,1,1), las=1)
plot(sppCount$year, sppCount$relCountLq, 
     type = "l", lwd = 3, 
     xlab = "Year", ylab = "Relative number of visits with observations", 
     ylim=c(0, max(sppCount$relCountLq)), 
     xaxp=c(2000, 2010, 10))
lines(sppCount$year, sppCount$relCountSd, 
      lwd=3, col="#78D2EB")
legend("bottomright", 
       legend=c("Libellula quadrimaculata","Sympetrum sanguineum"), 
       text.font = 3, col = c("black", "#78D2EB"), lwd = 3, bty = "n")
par(oldpar)
```
